Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-11-08 inference.tex
Typos.

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a fragment of FOL that satisfies the following syntactic properties:
\begin{itemize}
  \item All function symbols, modulo those that are drawn from
    $\Prop^0$ and $\Prop^1$, are uninterpreted and Boolean.
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.
\end{itemize}
The simple syntactic structure of the fragment already makes is
amenable for automatic reasoning via an off-the-shelf SMT solver, such
as Z3. The decidability of this fragment, however, is more subtle and
discussed below.

@@

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rule for $\bind$. Since an occurrence
of $\bind$ adds a quantifier to $|\nubar|$, if $s$ is a bind
expression nested inside a top-level bind expression (like
$\F_U(\stg)$ from the above example), then the prenex signature
of $\phi_s$ is $\forall^2\exists$.  Furthermore, if the subexpressions
of $s$ are neither {bind} nor $\existsl$ expressions, then none of the
$\phi_i$'s are quantified, and the prenex signature of $\phi = \bigwedge_i\phi_i \wedge \phi_s$
remains $\forall^2\exists$. A similar
observation holds when $s$ is an $\existsl$ expression nested inside a
top-level {bind} expression.  Since $\existsl$ is generated as a
result of stabilizing a SQL command transformer, which is always a
non-nested bind expression, the subexpression ($s'$) of $\existsl$ is
a non-nested bind expression. $s'$ is however nested inside a
top-level bind expression, hence its prenex signature is
$\forall^2\exists$.  Since $\existsl$ doesn't extend $\nubar$, the
prenex signature of $s$ remains $\forall^2\exists$. When $s$ is an
expression other than $\bind$ or $\existsl$, then $\phi_s$ is not a
quantified formula, and its prenex signature is trivially subsumed by
$\forall^2\exists$. Thus, for the subset of $\SL$, where bind
expressions are restricted to one level of nesting, the FOL formulas
generated by the encoding have the prenex signature as
$\forall^2\exists$.

>>>>

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a fragment of FOL that satisfies the following syntactic properties:
\begin{itemize}
  \item All function symbols, modulo those that are drawn from
    $\Prop^0$ and $\Prop^1$, are uninterpreted and boolean.
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.
\end{itemize}
The simple syntactic structure of the fragment already makes is
amenable for automatic reasoning via an off-the-shelf SMT solver, such
as Z3. The decidability of this fragment, however, is more subtle and
discussed below.

@@

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rule for $\bind$. Since an occurrence
of $\bind$ adds a quantifier to $|\nubar|$, if $s$ is a bind
expression nested inside a top-level bind expression (like
$\F_U(\stg)$ from the above example), then the prenex signature
of $\phi_s$ is $\forall^2\exists$.  Furthermore, if the subexpressions
of $s$ are neither {bind} nor $\existsl$ expressions, then none of the
$\phi_i$'s are quantified, and the prenex signature of $\phi = \bigwedge_i\phi_i \wedge \phi_s$
remains $\forall^2\exists$. A similar
observation holds when $s$ is an $\existsl$ expression nested inside a
top-level {bind} expression.  Since $\existsl$ is generated as a
result of stabilizing a SQL command transformer, which is always a
non-nested bind expression, the subexpression ($s'$) of $\existsl$ is
a non-nested bind expression. $s'$ is however nested inside a
top-level bind expression, hence its prenex signature is
$\forall^2\exists$.  Since $\existsl$ does not extend $\nubar$, the
prenex signature of $s$ remains $\forall^2\exists$. When $s$ is an
expression other than $\bind$ or $\existsl$, then $\phi_s$ is not a
quantified formula, and its prenex signature is trivially subsumed by
$\forall^2\exists$. Thus, for the subset of $\SL$, where bind
expressions are restricted to one level of nesting, the FOL formulas
generated by the encoding have the prenex signature as
$\forall^2\exists$.

### 2017-11-08 opsem.tex
Typos.

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), Integer and Boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\langle \bar{f}=\bar{e} \rangle$) constitute the
syntactic class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$),
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$, and executes at the top-level; our semantics does not
support nested transactions. The $\I$ in the \C{TXN} block syntax is
the transaction's isolation specification, whose purpose is explained
below.  Certain terms that only appear at run-time are also present in
$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
$\stg$) of records representing local and global database state, and a
runtime {\sf foreach} expression that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that the surface-level syntax of the \C{FOREACH}
command shown here is slightly different from the one used in previous
sections; its higher-order function has two arguments, $y$ and $z$,
which are invoked (during the reduction) with the set of
already-iterated items, and the current item, respectively. This form
of \C{FOREACH} lends itself to inductive reasoning that will be useful
for verification (Sec.~\ref{sec:reasoning}). Our language ensures
that all effectful actions are encapsulated within database commands,
and that all shared state among processes are only manipulated via
transactions and its supported operations.  In particular, we do not
consider programs in which objects resident on e.g., the OCaml heap
are concurrently manipulated by OCaml expressions as well as database
actions.

@@

The small-step semantics is stratified into a transaction-local
reduction relation, and a top-level reduction relation. The
transaction-local relation ($\stg \vdash (c,\stl) \stepsto
(c',\stl')$) defines a small-step reduction for a command inside a
transaction, when the database state is $\stg$; the command $c$
reduces to $c'$, while updating the transaction-local database $\stl$
to $\stl'$. The definition assumes a meta-function $\eval$ that
evaluates closed terms to values. The reduction relation for SQL
statements is defined straightforwardly. \C{INSERT} adds a new record
to $\stl$ after checking the uniqueness of its id. \C{DELETE} finds
the records in $\stg$ that match the search criteria defined by its
Boolean function argument, and adds the records to $\stl$ after
marking them for deletion. \C{SELECT} bounds the name introduced by
\C{LET} to the set of records from $\stg$ that match the search
criteria, and then executes the bound command $c$. \C{UPDATE} uses its
first function argument to compute the updated version of the records
that match the search criteria defined by its second function
argument. Updated records are added to $\stl$.

@@

The top-level reduction relation defines the small-step semantics of
transactions, and their parallel composition. A transaction comes
tagged with an \emph{isolation specification} $\I$, which has two components
$\I_e$ and $\I_c$, that dictate the timing and nature of interferences
that the transaction can witness, during its execution ($\I_e$), and
when it is about to commit ($\I_c$).  Formally, $\I_e$ and $\I_c$ are
predicates over the (current) transaction-local database state
($\stl$), the state ($\stg$) of the global database when the
transaction last took a step, and the current state ($\stg'$) of the
global database.  Intuitively, $\stg'\neq\stg$ indicates an
interference from another concurrent transaction, and the predicates
$\I_e$ and $\I_c$ decide if this interference is allowed or not,
taking into account the local database state ($\stl$). For instance,
as described in \S\ref{sec:motivation}, an SI transaction on
PostgreSQL defines $\I$ as follows:
\begin{smathpar}
\begin{array}{lcl}
\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
\end{array}
\end{smathpar}
This definition dictates that no change to the global database state
can be visible to an SI transaction while it executes ($\I_e$), and
there should be no concurrent updates to records written by the
transaction by other concurrently executing ones ($\I_c$).
To simplify the presentation, we use $\I$ instead of $\I_e$ and $\I_c$
when its destructed form is not required.

>>>>

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), integer and boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($\langle \bar{f}=\bar{e} \rangle$) constitute the
syntactic class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$),
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$, and executes at the top-level; our semantics does not
support nested transactions. The $\I$ in the \C{TXN} block syntax is
the transaction's isolation specification, whose purpose is explained
below.  Certain terms that only appear at run-time are also present in
$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
$\stg$) of records representing local and global database state, and a
runtime {\sf foreach} expression that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that the surface-level syntax of the \C{FOREACH}
command shown here is slightly different from the one used in previous
sections; its higher-order function has two arguments, $y$ and $z$,
which are invoked (during the reduction) with the set of
already-iterated items, and the current item, respectively. This form
of \C{FOREACH} lends itself to inductive reasoning that will be useful
for verification (Sec.~\ref{sec:reasoning}). Our language ensures
that all effectful actions are encapsulated within database commands,
and that all shared state among processes are only manipulated via
transactions and its supported operations.  In particular, we do not
consider programs in which objects resident on e.g., the OCaml heap
are concurrently manipulated by OCaml expressions as well as database
actions.

@@

The small-step semantics is stratified into a transaction-local
reduction relation, and a top-level reduction relation. The
transaction-local relation ($\stg \vdash (c,\stl) \stepsto
(c',\stl')$) defines a small-step reduction for a command inside a
transaction, when the database state is $\stg$; the command $c$
reduces to $c'$, while updating the transaction-local database $\stl$
to $\stl'$. The definition assumes a meta-function $\eval$ that
evaluates closed terms to values. The reduction relation for SQL
statements is defined straightforwardly. \C{INSERT} adds a new record
to $\stl$ after checking the uniqueness of its id. \C{DELETE} finds
the records in $\stg$ that match the search criteria defined by its
boolean function argument, and adds the records to $\stl$ after
marking them for deletion. \C{SELECT} bounds the name introduced by
\C{LET} to the set of records from $\stg$ that match the search
criteria, and then executes the bound command $c$. \C{UPDATE} uses its
first function argument to compute the updated version of the records
that match the search criteria defined by its second function
argument. Updated records are added to $\stl$.

@@

The top-level reduction relation defines the small-step semantics of
transactions, and their parallel composition. A transaction comes
tagged with an \emph{isolation specification} $\I$, which has two components
$\I_e$ and $\I_c$, that dictate the timing and nature of interferences
that the transaction can witness, during its execution ($\I_e$), and
when it is about to commit ($\I_c$).  Formally, $\I_e$ and $\I_c$ are
predicates over the (current) transaction-local database state
($\stl$), the state ($\stg$) of the global database when the
transaction last took a step, and the current state ($\stg'$) of the
global database.  Intuitively, $\stg'\neq\stg$ indicates an
interference from another concurrent transaction, and the predicates
$\I_e$ and $\I_c$ decide if this interference is allowed or not,
taking into account the local database state ($\stl$). For instance,
as described in \S\ref{sec:motivation}, an SI transaction on
PostgreSQL defines $\I$ as follows:
\begin{smathpar}
\begin{array}{lcl}
\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'
\end{array}
\end{smathpar}
This definition dictates that no change to the global database state
can be visible to an SI transaction while it executes ($\I_e$), and
there should be no concurrent updates to records written by the
transaction by other concurrently executing ones ($\I_c$).
To simplify the presentation, we use $\I$ instead of $\I_e$ and $\I_c$
when its destructed form is not required.

### 2017-11-08 reasoning.tex
Formatting fixed.
Typos.

%The
%stability of a binary assertion $Q$ w.r.t a ternary rely relation $\R$
%is defined as:
%\begin{smathpar}
%\begin{array}{c}
%\forall \stl,\stg,\stg'.~ Q(\stl,\stg) \conj \R(\stl,\stg,\stg')
%\Rightarrow Q(\stl,\stg')
%\end{array}
%\end{smathpar}
%That is, if $Q$ relates $\stl$ to $\stg$, and an interference allowed
%by the isolation specification (which implicitly considers the local
%state $\stl$) takes $\stg$ to $\stg'$, then $Q$ must also relate $\stl$
%to $\stg'$.

For the guarantee $G$ of a transaction to be valid, it must follow
from the post-condition $Q$ of the body, provided that $Q$ is stable
w.r.t the commit-time interference captured by $\R_c$. $\R_c$, like
$\R_e$, is computed as a rely relation modulo isolation, except that
commit-time isolation ($\I_c$) is considered. The validity of
$G$ is captured by the following implication:
\begin{smathpar}
\begin{array}{c}
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \rhd \stg)\spc
\end{array}
\end{smathpar}
In other words, if $Q$ relates the transaction-local database state
($\stl$) to the state of the global database ($\stg$) before a transaction
commits, then $G$ must relate the states of the global database before
and after the commit. The act of commit is captured by the flush
action ($\stl\rhd\stg$). Once we establish the validity of $G$ as a
faithful representative of the transaction, we can verify that the
transaction preserves the high-level invariant $I$ by checking the
stability of $I$ w.r.t $G$, i.e., $\forall \stg,\stg'.~I(\stg) \wedge
G(\stg,\stg') \Rightarrow I(\stg')$.

@@

\begin{definition}[\bfseries Semantics of RG judgments]
\label{def:rg-semantics}
The semantics of the global and transaction-local RG judgments are
defined thus:
\begin{smathpar}
\begin{array}{lclr}
\R \vdash \hoare{P}{c}{Q} & \defeq & \forall
  \stl,\stl',\stg,\stg',\nubar.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
  (\tbox{\cskip}_i, \stl',\stg')
  \Rightarrow Q(\stl',\stg') & \nubar = \texttt{freevars}(c)\\
\rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
  \Rightarrow (\forall \stg'.\; (c,\stg) 
    \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\
&&\hspace*{0.6in}\conj \texttt{step-guaranteed}(R,G,c,\stg)\\
\end{array}
\end{smathpar}

\noindent The
$\texttt{step-guaranteed}$ predicate used in the semantics of the
global RG judgment is defined below:\vspace*{-10pt}

\begin{smathpar}
\begin{array}{lcl}
\texttt{step-guaranteed}(R,G,c,\stg) &\defeq& \forall c',c''\stg',\stg''.
(c,\stg) \rstepssto{n} (c',\stg') \conj (c',\stg') \stepsto
  (c'',\stg'') \Rightarrow G(\stg',\stg'')\\
\end{array}
\end{smathpar}
\end{definition}

>>>>

%The
%stability of a binary assertion $Q$ w.r.t a ternary rely relation $\R$
%is defined as:
%\begin{smathpar}
%\begin{array}{c}
%\forall \stl,\stg,\stg'.~ Q(\stl,\stg) \conj \R(\stl,\stg,\stg')
%\Rightarrow Q(\stl,\stg')
%\end{array}
%\end{smathpar}
%That is, if $Q$ relates $\stl$ to $\stg$, and an interference allowed
%by the isolation specification (which implicitly considers the local
%state $\stl$) takes $\stg$ to $\stg'$, then $Q$ must also relate $\stl$
%to $\stg'$.
\raggedbottom
For the guarantee $G$ of a transaction to be valid, it must follow
from the post-condition $Q$ of the body, provided that $Q$ is stable
w.r.t the commit-time interference captured by $\R_c$. $\R_c$, like
$\R_e$, is computed as a rely relation modulo isolation, except that
commit-time isolation ($\I_c$) is considered. The validity of
$G$ is captured by the following implication:
\begin{smathpar}
\begin{array}{c}
  \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow G(\stg, \stl \rhd \stg)\spc
\end{array}
\end{smathpar}
In other words, if $Q$ relates the transaction-local database state
($\stl$) to the state of the global database ($\stg$) before a transaction
commits, then $G$ must relate the states of the global database before
and after the commit. The act of commit is captured by the flush
action ($\stl\rhd\stg$). Once we establish the validity of $G$ as a
faithful representative of the transaction, we can verify that the
transaction preserves the high-level invariant $I$ by checking the
stability of $I$ w.r.t $G$, i.e., $\forall \stg,\stg'.~I(\stg) \wedge
G(\stg,\stg') \Rightarrow I(\stg')$.

@@

\begin{definition}[\bfseries Semantics of RG judgments]
\label{def:rg-semantics}
The semantics of the global and transaction-local RG judgments are
defined thus:
\begin{smathpar}
\begin{array}{lclr}
\R \vdash \hoare{P}{c}{Q} & \defeq & \forall
  \stl,\stl',\stg,\stg'.~ P(\stl,\stg) \conj (\tbox{c}_i,\stl,\stg) \rstepssto{n}
  (\tbox{\cskip}_i, \stl',\stg')
  \Rightarrow Q(\stl',\stg') &\\
\rg{I,R}{c}{G,I} & \defeq &  \forall \stg.\, I(\stg)
  \Rightarrow (\forall \stg'.\; (c,\stg) 
    \rstepssto{n} (\cskip,\stg') \Rightarrow I(\stg')) \\
&&\hspace*{0.6in}\conj \texttt{txn-guaranteed}(R,G,c,\stg)\\
\end{array}
\end{smathpar}

\noindent The
$\texttt{txn-guaranteed}$ predicate used in the semantics of the
global RG judgment is defined below:\vspace*{-10pt}

\begin{smathpar}
\begin{array}{lcl}
\texttt{txn-guaranteed}(R,G,c,\stg) &\defeq& \forall c',c''\stg',\stg''.
(c,\stg) \rstepssto{n} (c',\stg') \conj (c',\stg') \stepsto
  (c'',\stg'') \Rightarrow G(\stg',\stg'')\\
\end{array}
\end{smathpar}
\end{definition}
